Visualizing demographic evolution using geographically inconsistent census data

### 2018-08-13 paper/method.tex
one past until LA

\begin{figure}
    \centering 
    \includegraphics[width=0.95\linewidth]{overview.pdf}
    \caption{Overview of the proposed method. A graph is generated combining the
        original census data, encoding the changing geographical information.
        The graph is partitioned into an hierarchy~\cite{markus2017}. The
        characteristics and evolution of the clusters are then visually
        represented.
        \label{fig:overview}}
\end{figure}

@@

\subsection{User interface}
The initial interface is illustrated in Figure~\ref{fig:ui}. Since demographic
data can be nuanced, with intricate interconnections, we decided against
presenting the interface using synthetic data, considering instead data from the
Chicago region between 1970 and 2010, using previous published studies as
corroboration. This region is known for its entrenched racial divide and the
emergence of a \emph{'young urban'} population with a higher education
level~\cite{Delmelle2016,Delmelle2017}. More details about this scenario are
presented in Section~\ref{sec:study}.

@@

The configuration panel, on top left in Figure~\ref{fig:ui}, displays which
aspects were considered and their weights (following Equation~\ref{eq:dist}). It
also includes other configuration options that can be altered without
re-processing the data, such as the number of clusters and the color option. The
gear button allows access to the other configuration options that do require
further processing, such as changing location, aspects, and weights. This panel
also includes the configuration of the selection mode for the trajectories,
which allows the user to set, add, or remove the next selected trajectories to
the current selection. This feature enables the analysis of complex sets of
trajectories.

@@

We adopt an \emph{enhanced} version of the traditional boxplot, which includes
the minimum, maximum, 25\% and 75\% quantiles for the current cluster, but also
the IQRs for the other clusters, in slightly larger and faded black rectangles.
We also color the current IQR according to its relevance. While there might be
some degree of similarity between the color schema for relevance and for
trajectory identification, none of the experts consulted reported confusion.
These simple changes allow the user to easily understand the composition of the
cluster and how it relates to the others. Violin plots~\cite{hintze1998violin}
could also be used, but with increased potential for cluttering the
representation of the other clusters, hindering the comparisons.

For instance, the boxplot that summarizes the purple cluster, detailed on
Figure~\ref{fig:boxplot}, illustrates that this cluster is best defined by the
proportion of the population with four or more years of college. The user can
quickly see that this is relevant because the corresponding IQR is colored with
the highest relevance present in the legend. It is also clear that, while this
cluster includes CTs that have between 10\% to 90\% of people in this variable,
approximately, half of them have about 60\% of the population with four or more
years of college. Since all the other IQRs are well separated, this is a
defining characteristic of this cluster. By clicking on the colored bar above
the boxplot, the user can select all trajectories that contain this cluster at
any point in time.

>>>>

\begin{figure}
    \centering 
    \includegraphics[width=0.975\linewidth]{overview.pdf}
    \caption{Overview of the proposed method. A graph is generated combining the
        original census data, encoding the changing geographical information.
        The graph is partitioned into an hierarchy~\cite{markus2017}. The
        characteristics and evolution of the clusters are then visually
        represented.
        \label{fig:overview}}
\end{figure}

@@

\subsection{User interface}
\label{sec:ui}
The initial interface is illustrated in Figure~\ref{fig:ui}. Since demographic
data can be nuanced, with intricate interconnections, we decided against
validating the interface using a synthetic dataset, considering instead data
from the Chicago region between 1970 and 2010, using previous published studies
as corroboration. This region is known for its entrenched racial divide and the
emergence of a \emph{'young urban'} population with a higher education
level~\cite{Delmelle2016,Delmelle2017}. More details about this dataset are
presented in Section~\ref{sec:study}, with an workflow example in
Figure~\ref{fig:chiWorkflow}.

@@

The configuration panel, on top left in Figure~\ref{fig:ui}, displays which
aspects were used and their weights (following Equation~\ref{eq:dist}). It also
includes other configuration options that can be altered without re-processing
the data, such as the number of clusters and the color option. The gear button
allows access to the other configuration options that do require further
processing, such as changing location, aspects, and weights. This panel also
includes the configuration of the selection mode for the trajectories, which
allows the user to set, add, or remove the next selected trajectories to the
current selection. This feature enables the analysis of complex sets of
trajectories.

@@

We adopt an \emph{enhanced} version of the traditional boxplot, which includes
the minimum, maximum, 25\% and 75\% quantiles for the current cluster, but also
the IQRs for the other clusters, in slightly larger and faded black rectangles.
We also color the current IQR according to its relevance. While there might be
some degree of similarity between the color schema for relevance and for
trajectory identification, none of the experts consulted reported confusion.
Indeed, one expert reported confusion regarding the grey rectangles that
represent the IQRs for other distributions when they are not colored; when that
variable is not relevant to the characterization of the cluster. These simple
changes allow the user to easily understand the composition of the cluster and
how it relates to the others. Violin plots~\cite{hintze1998violin} could also be
used, providing more information about the shape of the distribution, but with
increased potential for obstructing the representation of the other clusters.

For instance, the boxplot that summarizes the purple cluster illustrated in
Figure~\ref{fig:ui}, detailed on Figure~\ref{fig:boxplot}, illustrates that this
cluster is best defined by the proportion of the population with four or more
years of college. The user can quickly see that this is relevant because the
corresponding IQR is colored with the highest relevance present in the legend.
It is also clear that, while this cluster includes CTs that have between 10\% to
90\% of people in this variable, approximately, half of them have about 60\% of
the population with four or more years of college. Since all the other IQRs are
well separated, this is a defining characteristic of this cluster. Conversely,
the proportion of the population with one to three years of college is not
relevant, as indicated by black fill in the rectangle representing the IQR of
this cluster, in overlapping position with the rectangles of the other clusters.
By clicking on the colored bar above the boxplot, the user can select all
trajectories that contain this cluster at any point in time.

### 2018-08-13 paper/study.tex
one past until LA

We used five aspects for the USA: Education level, Family income, Marital
status, Occupation, and Race; and seven for Canada: Age, Education level, Home
language, Household Income, Marital status, Occupation, Place of birth, and
Religion.  While our method does not require geographic harmonization, it
requires matching variables over time. The supplementary material contains the
details of which census columns were used for each aspect. Income is slightly
inaccurate, even though we did correct for inflation. We grouped the original ranges into
three larger ranges, but they do not match precisely.

% \subsection{Chicago}
% \begin{figure*}
%     \centering
%  \includegraphics[width=0.85\linewidth]{newTeaser.pdf}
%  \caption{Boxplot representation for some of the clusters found, characterized
%  by White population (orange), high education (purple), and Black population
%  (green). The trajectories plot represents the associated population, showing an
%  increase in the purple cluster. The plot on the right compares the intra
%  quantile ranges for \emph{Four or more years of College}. While there was an
%  increase in the whole city (gray band), it was sharper in this
%  region.\label{fig:teaser}}
% \end{figure*}

\subsection{Los Angeles}
We selected a region around the metropolitan area of Los Angeles (LA). This
region follows urban density, not administrative boundaries. The summary of the
results using all aspects with equal weights and eight clusters is illustrated
in Figure~\ref{fig:la}, while the statistical description of the clusters is
illustrated in Figure~\ref{fig:laAll}, where the most relevant aspect of each
cluster is highlighted. From the trajectories plot, we can see that there is a
large but shrinking cluster, depicted in green, one increasing cluster in
purple, an almost constant orange cluster, a smaller but increasing pink
cluster, and three other small clusters. The corresponding map illustrates where
these clusters are located, and that they are somewhat geographically stable,
with some movement between the green, orange, and purple clusters.

@@

We considered a region that is approximately the administrative border of the
current city of Toronto and all seven available aspects. The results are
summarized in Figure~\ref{fig:to}, considering eight clusters.

\begin{figure}
    \centering 
    \includegraphics[width=0.95\linewidth]{to.pdf}
    \caption{Results for Toronto, with eight clusters.\label{fig:to}}
\end{figure}

The population with low percentage of University degrees is represented in
orange, mostly anglophone population in green, Asian immigrants in yellow, high
percentage of income in the highest bracket in purple, high percentage of Jewish
people in light green and brown, high percentage of Eastern Non-Christian
religion in pink, and high concentration of single people in dark gray.

From the trajectories plot, we can see that Toronto is more dynamic than both
Chicago and LA, with one cluster constantly shrinking, and others growing more
prominent. In the 1970s, the city was divided into four clusters: low number of
university degrees, Jewish population, majority anglophones, and high income,
with the latter being rather small. There is some variance in the other
characteristics of the orange cluster, but none as significant as the education
level, especially when compared to the other years. While the orange cluster
persists until 2011, the other clusters present have similar education profiles
to one another, leading the clustering method to use other aspects to
characterize them.

Toronto's dynamism also appears in that we see the influx of Asian population
starting in the 1980s and building thereafter, as represented by the yellow and
pink clusters. While both include a high percentage of people born in Asia, the
pink region is more defined by religion, with low percentage of university
degrees, and contains the lowest percentage  of people in the highest income
bracket for these clusters; the yellow is less defined by religion, and has
higher education and income, geographically corresponding to the Markham region,
know for its Chinese population. A similar division also happens for the two
Jewish clusters, where the light green cluster has lower education and income
levels than the brown cluster. The purple cluster of high income is somewhat
stable. This cluster includes until 2011 the Bridle Path neighborhood, known for
its wealthy population.  In 2011, however, it was classified into the yellow
cluster that includes Asian immigration, since about 35\% of the population for
this CT were born in Asia.  It maintained its income profile, with 85\% with 90k
CAD or more.

The grey regions of the map are another indication of Toronto's volatility over
this period. While there is a cluster that is painted dark grey, annotated by a
red circle in Figure~\ref{fig:to}, the light gray regions on the north part of the
city indicate volatility. Here the same region was classified into three or more
clusters for this five census period. Using the 'Add' mode for the trajectory
selection, we select these volatile regions. A subset of the details is
illustrated in Figure~\ref{fig:toVol}.

These regions account for about 5\% of Toronto's population. The whole region
was classified into the orange cluster in 1971 (low level of university degrees).
By 1991, most of the region was classified into the green cluster, representing
anglophone population, mostly Canadian born, with a higher level of education.
As the corresponding plot indicates, this trend in increasing education is
city-wide, but this region has people with better education than most.

In 2001, the purple cluster of high income annexes neighboring parts of the
volatile region, and the Asian born population increases sharply, as illustrated
by the appearance of the yellow cluster.  This cluster, once again, indicates
well educated, higher income, and about 30\%-50\% Asian born population. By
2011, the yellow cluster increased considerably, annexing parts of the high
income purple cluster, including the neighboring Bridle Path area.

>>>>

We used five aspects for the USA: Education level, Family income, Marital
status, Occupation, and Race; and seven for Canada: Age, Education level, Home
language, Household Income, Marital status, Occupation, Place of birth, and
Religion.  While our method does not require geographic harmonization, it
requires matching variables over time. The supplementary material contains the
details of which census columns were used for each aspect. Income is slightly
inaccurate, even though we did correct for official inflation. We grouped the
original ranges into three larger ranges, but they do not match precisely.

\subsection{Chicago}
We selected a region loosely following the administrative borders. The
demographic composition is well explored in the literature, with reports of
racial divide and gentrification~\cite{Delmelle2016,Delmelle2017,Hwang2014}.
While the definition of gentrification is still unclear and out of the scope of
this paper, we associate gentrification with higher education and income levels.

The initial state of the prototype is illustrated in Figure~\ref{fig:ui}, and
its findings are explained in Section~\ref{sec:ui}, where the racial divide is
clear. Starting from this initial state, the specific workflow used to identify
the existence and details of the gentrification process are illustrated in
Figure~\ref{fig:chiWorkflow}. For the users, the first step is to identify the
compositions of each cluster from the boxplots, so orange is associated with
majority of White population, green with majority Black, and purple with higher
proportion of four years of college or more (high education level). The expanded
version of the boxplots for the purple cluster shows a higher income level and
majority of occupations in administrative jobs, therefore the purple cluster
identifies gentrified regions.

The trajectories plot illustrates the process of gentrification, progressively
absorbing regions from the orange cluster (White). This corroborates results
from the literature reporting that Black neighborhoods are less likely to
gentrify~\cite{Hwang2014}. Moreover, this process is unlikely to be reversed, as
indicated by the limited number of trajectories leaving the purple stream. Next,
we select the region that is gentrified in 2010, by clicking on the
corresponding rectangle in the trajectories plot, updating the information on
the maps and the details portion of the interface.

The corresponding CTs are highlighted in the maps, where the spatial pattern is
clear, corresponding exactly to previous findings in the
literature~\cite{Hwang2014}. Further, we can also identify the regions that
gentrified earlier, with an stronger purple hue, compared to newer regions,
where the orange and green colors are still present. This also indicates from
which clusters they belonged before gentrifying. In the details portion of the
interface, the order of the aspects was updated to reflect the order of
relevance considering only the selected region. The most relevant aspect is the
Education, specifically "Four or more years of college", illustrated in the
rightmost portion of Figure~\ref{fig:chiWorkflow}, which is increasing for the
whole city (grey band), but faster and to a higher level in this region (black
band). Indeed, while the IQR for the city goes from 11\% to 45\%, the IQR for
this region goes from 55\% to 77\%. However, there are portions of this region
with significantly lower or higher proportions, as indicated by the dotted black
lines representing the minimum and maximum for the selected region.

\begin{figure*}
    \centering
 \includegraphics[width=0.85\linewidth]{newTeaser.pdf}
 \caption{Worflow to discover gentrification in Chicago: the purple cluster
 corresponds to high education / income. Its population is increasing over time,
 absorbing from the majority White cluster (orange). By selecting the purple
 cluster in 2010, the region is highlighted in the maps. The proportion of
 people with 4+ years of college is increasing in the whole city (grey IQRs),
 but significantly more in this region (black).\label{fig:chiWorkflow}}
\end{figure*}

We considered a region that is approximately the administrative border of the
current city of Toronto and all seven available aspects with equal weights. The
results are summarized in Figure~\ref{fig:to}, considering eight clusters.

\begin{figure}
    \centering 
    \includegraphics[width=0.95\linewidth]{to.pdf}
    \caption{Clustering results for Toronto, with eight clusters, including
    clusters representing Jewish population, high and low income, low education,
    and Asian immigration.\label{fig:to}}
\end{figure}

The population with low percentage of University degrees is represented in
orange, mostly anglophone population in green, Asian immigrants in yellow, high
percentage of income in the highest bracket in purple, high percentage of Jewish
people in light green and brown, high percentage of Eastern Non-Christian
religion in pink, and high concentration of single people in dark gray. From the
trajectories plot, we can see that Toronto is more dynamic than Chicago, with
one cluster constantly shrinking. In the 1970s, the city was divided into four
clusters: low number of university degrees, Jewish population, majority
anglophones, and high income. Interestingly, the more recent clusters that
absorbed regions from the orange cluster have similar education profiles and are
differentiated by other aspects. In this sense, the city is growing diverse,
changing from a common low education profile to a higher level of education with
more diversity in religion (pink) and immigration (yellow).

Indeed, the influx of Asian population is visible starting in the 1980s and
building thereafter, leading to the yellow and pink clusters. While both include
a high percentage of people born in Asia, the pink is more defined by religion,
with low percentage of university degrees, and contains the lowest percentage of
people in the highest income bracket for these clusters; the yellow is less
defined by religion, and has higher education and income, geographically
corresponding to the Markham region, know for its Chinese population. A similar
division also happens for the two Jewish clusters, where the light green cluster
has lower education and income levels than the brown cluster. The purple cluster
of high income is somewhat stable. This cluster includes the Bridle Path
neighborhood, known for its wealthy population, until 2011. In 2011 it was
classified into the yellow cluster of Asian immigration, since about 35\% of the
population for this CT were then born in Asia. The income distribution did not
change, with 85\% of the population with an income of 90k CAD or more.

The most significant indicator of Toronto's dynamism is the presence of grey
regions on the simplified color map; representing regions associated to three or
more clusters over this five census period. Using the 'Add' mode for the
trajectory selection, we select their trajectories, and a subset of the details
is illustrated in Figure~\ref{fig:toVol}. These regions account for about 5\% of
Toronto's population. The whole region was classified into the orange cluster in
1971 (low level of university degrees). By 1991, most of the region was
classified into the green cluster, representing anglophone population, mostly
Canadian born, with a higher level of education. As the corresponding plot
indicates, this trend in increasing education is city-wide, but this region has
people with better education than most.

In 2001, the purple cluster of high income annexes neighboring parts of the
volatile region, and the Asian born population increases sharply, as illustrated
by the appearance of the yellow cluster.  This cluster indicates well educated,
higher income, and about 30\%-50\% Asian born population. By 2011, the yellow
cluster increased considerably, annexing parts of the high income purple
cluster, including the neighboring Bridle Path area.

\subsection{Los Angeles}%TODO HERE
We selected a region around the metropolitan area of Los Angeles (LA), following
urban density, not administrative boundaries. The summary of the results using
all aspects with equal weights and eight clusters is illustrated in
Figure~\ref{fig:la}. The full statistical description of the clusters is
illustrated in Figure~\ref{fig:laAll}, where the most relevant aspect of each
cluster is highlighted. From the trajectories plot, we can see that there is a
large but shrinking cluster, depicted in green, one increasing cluster in
purple, an almost constant orange cluster, a smaller but increasing pink
cluster, and three other small clusters. The corresponding map illustrates where
these clusters are located, and that they are somewhat geographically stable,
with some movement between the green, orange, and purple clusters.

@@
